---
title: "Momentum"
format: html
editor: visual
---

```{r}
# rm(list = ls())
library(data.table)
library(PerformanceAnalytics)
library(roll)
library(ggplot2)
library(patchwork)
library(foreach)
```

```{r}
#| cache: true

# Import data
dt = fread("F:/strategies/momentum/prices.csv")

# Create yearmonth column
dt[, month := data.table::yearmon(date)]

# Sort
setorder(dt, symbol, date)
```

```{r}
# Upsample to monthly data
dt_month = dt[, .(close = last(close), 
                  beta = last(beta),
                  liquid_1000 = last(liquid_1000),
                  liquid_500 = last(liquid_500),
                  liquid_200 = last(liquid_200),
                  liquid_100 = last(liquid_100),
                  beta_rank_largest_90 = last(beta_rank_largest_90),
                  beta_rank_largest_95 = last(beta_rank_largest_95),
                  beta_rank_largest_99 = last(beta_rank_largest_99),
                  regime_spy_up_44 = last(regime_spy_up_44),
                  regime_spy_up_66 = last(regime_spy_up_66),
                  regime_spy_up_125 = last(regime_spy_up_125),
                  regime_spy_up_252 = last(regime_spy_up_252),
                  momentum_3 = last(momentum_3),
                  momentum_6 = last(momentum_6),
                  momentum_9 = last(momentum_9),
                  momentum_12 = last(momentum_12),
                  dollar_volume_zscore_winsorized3 = last(dollar_volume_zscore_winsorized3),
                  dollar_volume_zscore_winsorized6 = last(dollar_volume_zscore_winsorized6),
                  dollar_volume_zscore_winsorized9 = last(dollar_volume_zscore_winsorized9),
                  dollar_volume_zscore_winsorized12 = last(dollar_volume_zscore_winsorized12),
                  discreteness_3 = last(discreteness_3),
                  discreteness_6 = last(discreteness_6),
                  discreteness_9 = last(discreteness_9),
                  discreteness_12 = last(discreteness_12)
                  ), 
              by = .(symbol, month)]

# Calculate next month return
dt_month[, return := shift(close, -1, typ = "shift") / close - 1, by = symbol]

# Define ETF symbols we want to potentially invest in
symbols = c(
  'bal', 'corn', 'wood', 'cow', 'cow', 'jjg', 'soyb', 'sgg', 'weat', 'soyb', 
  'nib',  'jo', 'soyb', 'weat', 'uso', 'uhn', 'bno', 'uso', 'ung', 'uga', 
  'gld', 'cper', 'pall', 'slv', 'fxa', 'fxb', 'fxe', 'fxa', 'fxf', 'fxc', 
  'fxm', 'uup', 'qqq', 'ewj', 'spy', 'iwm', 'dia', 'vxx', 'eem', 'vwo', "tlt"
)
dt_month_etf = dt_month[symbol %in% symbols]

```

# Bin graphs

## Mean returns across bins

```{r}
# Show mean and meadian return for bins across date and momentum_12
ret_bins = copy(dt_month_etf)
ret_bins[, bin := cut(momentum_12, breaks = quantile(momentum_12, probs = seq(0, 1, 0.1), na.rm = TRUE), 
                      include.lowest = TRUE, labels = FALSE)]
ret_bins = ret_bins[, .(mean_return = mean(return, na.rm = TRUE),
                        median_return = median(return, na.rm = TRUE)), by = .(bin)]
ret_bins = na.omit(ret_bins)
ggplot(ret_bins, aes(x = bin, y = mean_return)) + 
  geom_bar(stat = "identity")
```

### Media nreturns across bins

```{r}
ggplot(ret_bins, aes(x = bin, y = median_return)) + 
  geom_bar(stat = "identity")
```

### Mean returns across bins for 500 most liquid

```{r}
ret_bins = dt_month[liquid_500 == TRUE]
ret_bins[, bin := cut(momentum_12, breaks = quantile(momentum_12, probs = seq(0, 1, 0.1), na.rm = TRUE), 
                      include.lowest = TRUE, labels = FALSE)]
ret_bins = ret_bins[, .(mean_return = mean(return, na.rm = TRUE),
                        median_return = median(return, na.rm = TRUE)), by = .(bin)]
ret_bins = na.omit(ret_bins)
ggplot(ret_bins, aes(x = bin, y = mean_return)) + 
  geom_bar(stat = "identity")
```

### Median returns across bins for 500 most liquid

```{r}
ggplot(ret_bins, aes(x = bin, y = median_return)) + 
  geom_bar(stat = "identity")
```

### Bins over time

```{r}
ret_bins = dt_month[liquid_500 == TRUE]
ret_bins[, bin := cut(momentum_12, breaks = quantile(momentum_12, probs = seq(0, 1, 0.1), na.rm = TRUE), 
                      include.lowest = TRUE, labels = FALSE)]
ret_bins = ret_bins[, .(mean_return = mean(return, na.rm = TRUE),
                        median_return = median(return, na.rm = TRUE)), 
                    by = .(bin, year = as.integer(month))]
ret_bins = na.omit(ret_bins)
ggplot(ret_bins, aes(x = bin, y = mean_return)) + 
  geom_bar(stat = "identity") +
  facet_wrap(~year, scales = "free_y")
```

```{r}
ggplot(ret_bins, aes(x = bin, y = median_return)) + 
  geom_bar(stat = "identity") +
  facet_wrap(~year, scales = "free_y")
```

# Equity curves

## 12 month momentum on whole unvierse

Assumptions: - We will keep 30 companies with highest momentum - We will rebalance monthly - We will equally weight all companies in the portfolio

```{r}
# Keep 30 companies with highest momentum
sample_ = copy(dt_month)
setorder(sample_, month, -momentum_12)
sample_ = na.omit(sample_, cols = c("momentum_12", "return"))[, first(.SD, 30), by = month]

# Calculate portfolio reutrns
sample_[, weights := 1 / length(close), by = month]
sample_[, all(weights == first(weights))] # test if all weights are the same
dt_portfolio = sample_[, .(portfolio_return = sum(weights * return)), by = month]
dt_xts = as.xts.data.table(dt_portfolio[, .(as.Date(zoo::as.yearmon(month)), 
                                            portfolio_return)])
charts.PerformanceSummary(dt_xts)
```

```{r}
charts.PerformanceSummary(dt_xts["1998/2007"], main = "1998/2007")
```

```{r}
charts.PerformanceSummary(dt_xts["2007/2018"], main = "2007/2018")
```

```{r}
charts.PerformanceSummary(dt_xts["2018/"], main = "2018/")
```

## 12 month momentum on 500 most liquid companies

```{r}
# Same analysis on universe of 500 most liquid companies
sample_ = copy(dt_month)
setorder(sample_, month, -momentum_12)
sample_ = na.omit(sample_, cols = c("momentum_12", "return"))
sample_ = sample_[liquid_1000 == TRUE]
sample_ = sample_[, first(.SD, 30), by = month]
sample_[, weights := 1 / length(close), by = month]
sample_[, all(weights == first(weights))] # test if all weights are the same
dt_portfolio = sample_[, .(portfolio_return = sum(weights * return)), by = month]
dt_xts = as.xts.data.table(dt_portfolio[, .(as.Date(zoo::as.yearmon(month)),
                                            portfolio_return)])
charts.PerformanceSummary(dt_xts, main = "500 most liquid companies")
```

## 12 month momentum on 500 most liquid companies

```{r}
# Same analysis on universe of 500 most liquid companies
sample_ = copy(dt_month)
setorder(sample_, month, -momentum_12)
sample_ = na.omit(sample_, cols = c("momentum_12", "return"))
sample_ = sample_[liquid_500 == TRUE]
sample_ = sample_[, first(.SD, 30), by = month]
sample_[, weights := 1 / length(close), by = month]
sample_[, all(weights == first(weights))] # test if all weights are the same
dt_portfolio = sample_[, .(portfolio_return = sum(weights * return)), by = month]
dt_xts = as.xts.data.table(dt_portfolio[, .(as.Date(zoo::as.yearmon(month)),
                                            portfolio_return)])
charts.PerformanceSummary(dt_xts, main = "500 most liquid companies")
```

```{r}
# Same analysis on universe of 100 most liquid companies
sample_ = copy(dt_month)
setorder(sample_, month, -momentum_12)
sample_ = na.omit(sample_, cols = c("momentum_12", "return"))
sample_ = sample_[liquid_200 == TRUE]
sample_ = sample_[, first(.SD, 30), by = month]
sample_[, weights := 1 / length(close), by = month]
sample_[, all(weights == first(weights))] # test if all weights are the same
dt_portfolio = sample_[, .(portfolio_return = sum(weights * return)), by = month]
dt_xts = as.xts.data.table(dt_portfolio[, .(as.Date(zoo::as.yearmon(month)),
                                            portfolio_return)])
charts.PerformanceSummary(dt_xts, main = "500 most liquid companies")
```

# Beta adjustment

```{r}
sample_ = copy(dt_month)
sample_ = sample_[beta_rank_largest_95 == 0]
sample_[, momentum_12 := momentum_12 / beta]
setorder(sample_, month, -momentum_12)
sample_ = na.omit(sample_, cols = c("momentum_12", "return"))
sample_ = sample_[liquid_1000 == TRUE]
sample_ = sample_[, first(.SD, 30), by = month]
sample_[, weights := 1 / length(close), by = month]
sample_[, all(weights == first(weights))] # test if all weights are the same
dt_portfolio = sample_[, .(portfolio_return = sum(weights * return)), by = month]
dt_xts = as.xts.data.table(dt_portfolio[, .(as.Date(zoo::as.yearmon(month)),
                                            portfolio_return)])
charts.PerformanceSummary(dt_xts, main = "500 most liquid companies")
```

# Regimes

```{r}
sample_ = copy(dt_month)
setorder(sample_, month, -momentum_12)
sample_ = na.omit(sample_, cols = c("momentum_12", "return"))
sample_ = sample_[liquid_1000 == TRUE]
sample_ = sample_[, first(.SD, 30), by = month]
sample_ = sample_[regime_spy_up_125 == 1]
sample_[, weights := 1 / length(close), by = month]
sample_[, all(weights == first(weights))] # test if all weights are the same
dt_portfolio = sample_[, .(portfolio_return = sum(weights * return)), by = month]
dt_xts = as.xts.data.table(dt_portfolio[, .(as.Date(zoo::as.yearmon(month)),
                                            portfolio_return)])
charts.PerformanceSummary(dt_xts, main = "500 most liquid companies")
```

# Linear regression of returns on momentum indicators

```{r}
# Dates
dates = seq.Date(as.Date("2023-10-01"), as.Date("2024-09-01"), by = "month")
dates = data.table::yearmon(dates)

# Rolling linear regression of returns on momenutm indicators
data_ = dt_month[month %in% dates]
data_[, month := round(month, 1)]
# remove outliers
data_ = data_[momentum_12 < quantile(momentum_12, 0.99, na.rm = TRUE)]
data_ = data_[momentum_12 > quantile(momentum_12, 0.01, na.rm = TRUE)]
data_ = data_[return < quantile(return, 0.99, na.rm = TRUE)]
data_ = data_[return > quantile(return, 0.01, na.rm = TRUE)]

# ggplot of linear regression of momentum on returns
ggplot(data_, aes(x = momentum_12, y = return)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  facet_wrap(~month, scales = "free_y")

```

# Try different parameters

```{r}
# Define parameters
coarse_universe = c("liquid_100", "liquid_200", "liquid_500", "liquid_1000",
                    "all")
beta_filtering = c("beta_rank_largest_90", "beta_rank_largest_95", 
                   "beta_rank_largest_99", "all")
beta_adjustment = c(TRUE, FALSE)
momentun_vars = colnames(dt_month)[grepl("momentum", colnames(dt_month))]
dvzw_thresholds = c(2:5, NA)
discreteness_log = c(TRUE, FALSE)
n = c(10, 20, 30, 50, 100)
regimes = c("regime_spy_up_44", "regime_spy_up_66", 
            "regime_spy_up_125", "regime_spy_up_252", "all")
params = expand.grid(coarse_universe = coarse_universe,
                     beta_filtering = beta_filtering,
                     beta_adjustment = beta_adjustment,
                     momentun_vars = momentun_vars,
                     dvzw_thresholds = dvzw_thresholds,
                     discreteness_log = discreteness_log,
                     n = n, 
                     regimes = regimes,
                     stringsAsFactors = FALSE)
setDT(params)
# head(params)
# nrow(params)
```

```{r}
# Calculate momentum for all coarse universes
cluster = parallel::makeCluster(12, type = "PSOCK")
doParallel::registerDoParallel(cl = cluster)
foreach::getDoParRegistered()
foreach::getDoParWorkers()

# TEST
# params = params[1:100]

start_time = Sys.time()
res = foreach(i=1:nrow(params), 
              .combine='rbind', 
              .packages = c("data.table", "xts", "PerformanceAnalytics")) %dopar% {
  # i = 4
  # print(i)
  # parameters
  coarse_ = params$coarse_universe[i]
  beta_filtering_ = params$beta_filtering[i]
  momentum_ = params$momentun_vars[i]
  dvzw_ = params$dvzw_thresholds[i]
  discreteness_ = params$discreteness_log[i]
  n_ = params$n[i]
  regimes_ = params$regimes[i]
  beta_adj_ = params$beta_adjustment[i]

  dt_mom = na.omit(dt_month_etf, cols = c(momentum_, "return"))
  if (coarse_ != "all") {
   dt_mom = dt_mom[get(coarse_) == TRUE] 
  }
  if (beta_filtering_ != "all") {
   dt_mom = dt_mom[get(beta_filtering_) == 0]
  }
  if (regimes_ != "all") {
   dt_mom = dt_mom[get(regimes_) == 1] 
  }
  if (!is.na(dvzw_)) {
   dt_mom = dt_mom[get(paste0("dollar_volume_zscore_winsorized", gsub("momentum_", "", momentum_))) < dvzw_]
  }
  if (beta_adj_) {
   dt_mom[, x := x / beta, env = list(x = momentum_)]
  }
  if (discreteness_) {
   dt_mom = dt_mom[get(paste0("discreteness_", gsub("momentum_", "", momentum_))) < 0]
  }
  setorderv(dt_mom, c("month", momentum_), order = c(1, -1))
  dt_mom = dt_mom[, first(.SD, n_), by = month]
  dt_mom[, weights := 1 / length(close), by = month]
  dt_mom[, all(weights == first(weights))] # test if all weights are the same
  dt_portfolio = dt_mom[, .(portfolio_return = sum(weights * return)), by = month]
  dt_xts = as.xts.data.table(dt_portfolio[, .(as.Date(zoo::as.yearmon(month)),
                                              portfolio_return)])
  cbind(
   annualized_return = Return.annualized(dt_xts),
   sharpe_ratio = suppressMessages(SharpeRatio(dt_xts)[1, 1]),
   max_dd = maxDrawdown(dt_xts)
  )
  # params[i, let(
  #  annualized_return = Return.annualized(dt_xts),
  #  sharpe_ratio = suppressMessages(SharpeRatio(dt_xts)[1, 1]),
  #  max_dd = maxDrawdown(dt_xts)
  # )]
}
end_time = Sys.time()
print(end_time - start_time)
# Time difference of 56.194 mins
# Stop cluster
parallel::stopCluster(cluster)

# Merge results and parameters
res = cbind.data.frame(params, res)
setDT(res)

# Order by sharpe ratio
setorder(res, -sharpe_ratio)
```

```{r}
# Save results
fwrite(res, "F:/strategies/momentum/results.csv")

# res = fread("F:/strategies/momentum/results.csv")

head(res, 20)
```

```{r}
head(res)
```

